Backpropagating Gravitational wave events

### 2022-05-13 src/ms.tex
adding in extra text

While forward modelling entire populations has been the canonical pathway to compare simulations and GW observations,
it is not necessarily the most efficient way to constrain the underlying physical processes, or "hyper-parameters", which drive the evolution of the population.
One of the most common skepticism of population synthesis studies is that these studies employ simulation sets which often 
cover a small subdomain of the entire parameter space of uncertain physics, or only vary one uncertain process per simulation
(e.g. common envelope efficiency) while holding all others fixed to some "default" values.
On top of this, these values are often not calibrated to data.
Even with recent developments in using emulators to speed up the simulation process, training the emulators 
still requires a significant amount of computation to cover a wide range of uncertainty.
As an example, to emulate a model with $10$ parameters, the simplest way to construct a training set of simulations is to 
run a grid which varies combinations of uncertain physics.
For 10 uncertain physical processes, if we consider $2$ variations for each physical process 
this would still require $1024$ simulations.
Because population synthesis simulations require hundreds to thousands of cpu hours,
training an emulator which spans a high dimensional uncertainty space remains an impossibility at present.

@@

In this paper, we propose a method to "backward" model each GW event to its progenitor state 
while allowing the hyper-parameters to vary across their full range of physical uncertainty
following a two-stage process.
First, we solve a root finding problem to obtain guesses that are likely to produce the desired system properties in the joint progenitor--hyper-parameter space.
We then sample the posterior in the joint space with Markov chain Monte Carlo (MCMC) algorithm that is initialized using the roots found in the previous stage.
The rest of the paper is structured as follows: We detail our method in Sec. ~\ref{sec:method}.
In Sec. ~\ref{sec:result}, we show the results of applying our method to a number of real events.
We discuss the implications of this work and future direction in Sec. ~\ref{sec:discussion}.

@@

We define the parameters that describe the progenitor's zero age main sequence (ZAMS) properties
masses and eccentricity as progenitor parameters $\bm{\theta'}$, the parameters
that control different uncertain physical prescriptions such as wind strength and common
envelope efficiency as hyper-parameters $\bm{\lambda}$, and random variables
that affect certain stochastic process such as whether CO natal kick unbinds a
binary as $\bm{X}$. To avoid clutter in the following derivation, we denote all
the parameters related to mapping a particular progenitor system into a GW event
collectively as evolutionary parameters $\bm{\Theta}$, which includes
$\bm{\theta'}$, $\bm{\lambda}$, and $\bm{X}$.

@@

% \kw{Talk about implication on single event posterior.}
Once we obtain a set of binaries which successfully map the ZAMS parameters and hyper-parameters to BBH merger masses, 
we re-evolve the set of ZAMS parameters with the same physical assumptions as A21, but vary the common envelope efficiency 
to explore how keeping a fixed model which only varies one hyper-parameter affects our analysis. 
Figure~\ref{fig:compare_fixed_variable} shows the distribution of ZAMS masses for three models (first through third columns), 
each with a different $\alpha$ but the same hyper-parameters as A21, 
as well as the results of our sampling which allows our hyper-parameters to vary (fourth column). We find that holding the accretion 
efficiency, common envelope and natal kick physics to fixed values greatly reduces the ZAMS parameter space 
that produces GW150914-like mergers. In contrast, by allowing the hyper-parameters to fully span the model uncertainty, we 
find that there are distinct ZAMS parameters which produce GW150914-like mergers and conversely, ZAMS parameters which 
fail to produce GW150914-like mergers \emph{regardless of our hyper-parameter choice.}
\kw{@Katie, can you put a frame around the fourth column highlighting it is this work?}

>>>>

While forward modelling entire populations has been the canonical pathway to compare simulations and GW observations,
it is not necessarily the most efficient way to constrain the underlying physical processes, or ``hyper-parameters", which drive the evolution of the population.
One of the most common skepticisms of population synthesis studies is that \kb{they} employ simulation sets which often 
cover a small subdomain of the entire parameter space of uncertain physics, or only vary one uncertain process per simulation
(e.g. common envelope efficiency) while holding all others fixed to some ``default" values.
On top of this, these values are often not calibrated to data.
Even with recent developments in using emulators to speed up the simulation process, training the emulators 
still requires a significant amount of computation to cover a wide range of uncertainty.
As an example, to emulate a model with $10$ parameters, the simplest way to construct a training set of simulations is to 
run a grid which varies combinations of uncertain physics.
For $10$ uncertain physical processes, if we consider $2$ variations for each physical process 
this would still require $1024$ simulations.
Because population synthesis simulations require hundreds to thousands of cpu hours,
training an emulator which spans a high dimensional uncertainty space remains an impossibility at present.

@@

In this paper, we propose a method to ``backward" model each GW event to its progenitor state 
while allowing the hyper-parameters to vary across their full range of physical uncertainty
following a two-stage process.
First, we solve a root finding problem to obtain guesses that are likely to produce the desired system properties in the joint progenitor--hyper-parameter space.
We then sample the posterior in the joint space with Markov chain Monte Carlo (MCMC) algorithm that is initialized using the roots found in the previous stage.
The rest of the paper is structured as follows: We detail our method in Sec. ~\ref{sec:method}.
In Sec. ~\ref{sec:result}, we show the results of applying our method to a number of real events.
We discuss the implications of this work and future direction in Sec. ~\ref{sec:discussion}.

@@

We define the progenitor's zero age main sequence (ZAMS) properties like
mass, orbital period, and eccentricity as progenitor parameters $\bm{\theta'}$, the parameters
that control different uncertain physical prescriptions such as wind strength and common
envelope efficiency as hyper-parameters $\bm{\lambda}$, and random variables
that affect certain stochastic process such as whether CO natal kick unbinds a
binary as $\bm{X}$. To avoid clutter in the following derivation, we denote all
the parameters related to mapping a particular progenitor system into a GW event
collectively as evolutionary parameters $\bm{\Theta}$, which includes
$\bm{\theta'}$, $\bm{\lambda}$, and $\bm{X}$.

@@

% \kw{Talk about implication on single event posterior.}
Once we obtain a set of binaries which successfully map the ZAMS parameters and hyper-parameters to BBH merger masses, 
we re-evolve the set of ZAMS parameters with the same physical assumptions as A21, but vary the common envelope efficiency 
to explore how keeping a fixed model which only varies one hyper-parameter affects our analysis. 
Figure~\ref{fig:compare_fixed_variable} shows the distribution of ZAMS masses for three models (first through third columns), 
each with a different $\alpha$ but the same hyper-parameters as A21, 
as well as the results of our sampling which allows our hyper-parameters to vary (fourth column). We find that holding the accretion 
efficiency, $f_{\rm{acc}}$, and common envelope ejection efficiency, $\alpha$, to fixed values greatly reduces the ZAMS parameter space 
that produces GW150914-like mergers. In contrast, by allowing the hyper-parameters to fully span the model uncertainty, we 
find that there are distinct ZAMS parameters which produce GW150914-like mergers and conversely, ZAMS parameters which 
fail to produce GW150914-like mergers \emph{regardless of our hyper-parameter choice.} For primary ZAMS masses between 
$\sim95-135\,M_{\odot}$ and secondary ZAMS masses between $\sim50-95\,M_{\odot}$, we find that there is no combination of accretion
efficiency and common envelope ejection efficiency which produces merging BBHs that have masses consistent with GW150914's posteriors.
In this region, if BBHs with masses consistent with GW150914's masses are produced through a combination of $\alpha$ and 
$f_{\rm{acc}}$, they do not merge in a Hubble time. Conversely, in the region of the fourth column of Figure~\ref{fig:compare_fixed_variable}
which does produce GW150914-like mergers, the combination of $\alpha$, $f_{\rm{acc}}$, and the ZAMS masses and orbital separations 
balance to produce BHs with the correct masses that merge within a Hubble time.
%\kw{@Katie, can you put a frame around the fourth column highlighting it is this work?}
%\kb{done!}
